---
title: "Assignment 2"
format: html
editor_options: 
  chunk_output_type: console
---

## Intro
```{r}
#| message: false  

# Libraries
library(tidyverse)

# file names
datadir_raw <- file.path("data","raw")

datadir_processed <- file.path("data", "processed/")

snowsurvey_file <- "snow_cover.csv"
```

```{r}
# Import processed data with fixed Snow_cover column
snowsurvey_csv <- read_csv(file.path(datadir_processed, snowsurvey_file))

# Look at Water_cover, Land_cover, and Total_cover
glimpse(snowsurvey_csv)
```

## Land_cover

### Data Exploration
```{r}
# Check out why it's a character variable
snowsurvey_csv %>%
 count(Land_cover) %>% 
  print(n = 101)
```
There are interesting results. Similar to the Snow_cover, there are periods, dashes, and unks, but there are also negative numbers now. 

```{r}
# Check on non-numeric values
snowsurvey_csv %>% 
  count(Land_cover) %>% 
  filter(is.na(as.numeric(Land_cover))) %>% 
  view()
```

### Data Cleaning

#### `.` values
```{r}
# Check observations where land_cover is a dot
snowsurvey_csv %>%
  filter(Land_cover == ".") %>%
  View()
```

I'm going to replace the dots with NAs to match what we did in the Snow_cover manipulation and because the metadata doesn't specify.
```{r}
snowsurvey_fixed <- snowsurvey_csv %>% 
  mutate(Land_cover = ifelse(Land_cover == ".", NA, Land_cover))
```

#### `-` values

Is the problem true for '-' as well?
```{r}
# Check observations where land_cover is a dash
snowsurvey_csv %>%
  filter(Land_cover == "-") %>%
  View() 
```


Yep, the problem is like the dots, so I'm also going to set this to NA. 
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover == "-", NA, Land_cover))

```
#### `n/a` values

"n/a" will also be replaced with NA, but I will view it first.

```{r}
snowsurvey_csv %>%
  filter(Land_cover == "n/a") %>%
  View()
  
```


Yep, it has the same problem.
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover == "n/a", NA, Land_cover))
 
```

#### `unk` values

'unk' is most likely short for unknown, so I will also replace this with NA. 

```{r}
snowsurvey_csv %>%  
  filter(Land_cover == "unk") %>%
  View()
```


```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover == "unk", NA, Land_cover))
```

### Going numeric

Check that there are no non-numerics left in the Land_cover column.
```{r}
snowsurvey_fixed %>%
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```

Now that there are only numerics left, I can change the column type to numeric
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = as.numeric(Land_cover))
```

### More cleaning
I will now check that the percentages are less than 100.
```{r}
 snowsurvey_fixed %>% 
  filter(Land_cover > 100)
```

Nothing is more than 100, so nothing there needs to be changed. But, I will check for negatives because those would also not make sense. 
```{r}
 snowsurvey_fixed %>% 
  filter(Land_cover < 0)
```
There are two values that are negatives, and one is negative 100. I'm going to set those both to NA. 

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover < 0, NA, Land_cover))

# Check Land_cover again
snowsurvey_fixed %>%
 count(Land_cover) %>% 
  print(n = 101)
```
Everything looks good, so I'm going to move on to the Water_cover column with the new fixed dataframe.

## Water_cover

### Data Exploration
```{r}
# Check out why it's a character variable
snowsurvey_fixed %>%
 count(Water_cover) %>% 
  print(n = 95)
```
There seem to be the same problems as with Land_cover.

```{r}
# Check on non-numeric values
snowsurvey_fixed %>% 
  count(Water_cover) %>% 
  filter(is.na(as.numeric(Water_cover))) %>% 
  view()
```

### Data Cleaning

#### `.` values
```{r}
# Check observations where water_cover is a dot
snowsurvey_fixed %>%
  filter(Water_cover == ".") %>%
  View()
```

I'm going to replace the dots with NAs to match what we did in the Snow_cover and Land_cover manipulation and because the metadata doesn't specify.
```{r}
snowsurvey_fixed <- snowsurvey_csv %>% 
  mutate(Water_cover = ifelse(Water_cover == ".", NA, Water_cover))
```

#### `-` values

Is the problem true for '-' as well?
```{r}
# Check observations where water_cover is a dash
snowsurvey_fixed %>%
  filter(Water_cover == "-") %>%
  View()
```


Yep, the problem is like the dots, so I'm also going to set this to NA. 
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover))

```
#### `n/a` values

"n/a" will also be replaced with NA, but I will view it first.

```{r}
snowsurvey_fixed %>%
  filter(Water_cover == "n/a") %>%
  View()
  
```


Yep, it has the same problem.
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover))
 
```

#### `unk` values

'unk' is most likely short for unknown, so I will also replace this with NA. 

```{r}
snowsurvey_fixed %>%  
  filter(Water_cover == "unk") %>%
  View()
```


```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover == "unk", NA, Water_cover))
```

### Going numeric

Check that there are no non-numerics left in the Water_cover column.
```{r}
snowsurvey_fixed %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```

Now that there are only numerics left, I can change the column type to numeric.
```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = as.numeric(Water_cover))
```

### More cleaning
I will now check that the percentages are less than 100.
```{r}
 snowsurvey_fixed %>% 
  filter(Water_cover > 100)
```

There is a value of 353 that does not make sense, and the metadata doesn't specify, so I will set to NA like we did with Snow_cover when there were values over 100.

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```

Let's check for negative values now.
```{r}
 snowsurvey_fixed %>% 
  filter(Water_cover < 0)
```
No negative values!

```{r}
# Check Water_cover again
snowsurvey_fixed %>%
 count(Water_cover) %>% 
  print(n = 74)
```
Everything looks good! Let's move to total_cover column.

## Total_cover

### Data Exploration
```{r}
# Look at total_cover
snowsurvey_fixed %>%
 count(Total_cover) %>% 
  print(n = 39)
```
Same problems as the last two, so I will quickly replace all the non-numeric to match what we did with Snow_cover, Land_cover, and Water_cover and change the column to numeric.
```{r}
# Check on non-numeric values
snowsurvey_fixed %>% 
  count(Total_cover) %>% 
  filter(is.na(as.numeric(Total_cover))) %>% 
  view()
```

### Data Cleaning

```{r}
# No mention of these non-numeric values in metadata so assume they are NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = ifelse(Total_cover == ".", NA, Total_cover)) %>% 
  mutate(Total_cover = ifelse(Total_cover == "-", NA, Total_cover)) %>% 
  mutate(Total_cover = ifelse(Total_cover == "n/a", NA, Total_cover)) %>% 
  mutate(Total_cover = ifelse(Total_cover == "<row r=\\64\\ spans=\\1:11\\ x14ac:dyDescent=\\0.2\\>", NA, Total_cover))

# Check to make sure NAs are the only non-numeric value
snowsurvey_fixed %>%
  count(Total_cover) %>% 
  filter(is.na(as.numeric(Total_cover)))

# Now change Total_cover to numeric column
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = as.numeric(Total_cover))

```


### More cleaning
Let's make sure total_cover isn't more than 100 and the sum of the other three columns (Snow_cover, Land_cover, Water_cover). 
```{r}
# View values in Total_cover
snowsurvey_fixed %>%
  count(Total_cover) %>% 
  print(n = 35)

# There are some values that are not 100
# Remake `Total_cover` column to make sure it's the sum of the other three columns
snowsurvey_fixed$Total_cover <- snowsurvey_fixed$Snow_cover + snowsurvey_fixed$Land_cover + snowsurvey_fixed$Water_cover

# Check for values less than 100
snowsurvey_fixed %>% filter(Total_cover < 100)

# The metadata only says this column should always equal 100, so I will replace any non 100s to NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = ifelse(Total_cover < 100, NA, Total_cover))

# Check for values over 100
snowsurvey_fixed %>% filter(Total_cover > 100)

# The metadata only says this column should always equal 100, so change non 100s to NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = ifelse(Total_cover > 100, NA, Total_cover))

# Check to make sure there are only vlaues of 100 or NA in Total_cover
snowsurvey_fixed %>%
  count(Total_cover)
```
Now that Total_cover is numeric and either 100 or NA, it looks good!

### Save cleaned df to csv
```{r}
# Write the file
write_csv(snowsurvey_fixed, file.path("data", "processed", "all_cover_fixed_kimberleewong.csv"))
```

